\textbf{Dynamic Hover Detection:\\Creating GUIs in Elm}

### 2013-12-15 paper/paper.tex
Decapitalize Element, Boolean, when not in code

We can print the current area to the screen with \texttt{main = lift asText
area}. The primitive \texttt{asText : a -> Element} renders almost
anything into an Element, which represents a DOM element.\footnote{Those
    following along in an Elm compiler, such as the one available at
    \texttt{elm-lang.org/try}, should add \texttt{import Window} to the top of
the file.}

@@

Elm's \texttt{Graphics.Input} library provides a function to obtain the hover
information from an Element.

@@

The returned Element is visually identical to the argument, but now detects
hover information. The signal of Booleans reflects the hover status of
the this Element, not the original. This function works well when the
Element is pure (not a signal). For example, if the top-level menu items are
known statically and never change, they can be rendered as pure Elements with
this function. We can also create a function, perhaps of type \texttt{Bool ->
Element}, which we can lift onto the hover data to display the submenu only on
mouseover. The result is a value of type \texttt{Signal Element}. If we try to
lift \texttt{hoverable} to accept this value, we get

@@

This is rather unfortunate, as we need to detect hover information on a dynamic
element for a number of reasons. First, after a menu is extended by hovering
over its parent, it must remain so as long as the mouse hovers over the menu
itself. Without this information, the menu will disappear as soon as we try to
mouse over it. Secondly, if we wish the indicate the currently hovered-over item
to the user by visually highlighting it, the element must be dynamic. Thirdly,
we do not wish to require that menu labels be string literals in the source
code, which would be necessary if we could not use signals of elements. Rather,
they may be retrieved or generated in some other manner, or they may change as
the user switches among applications. Finally, in order to determine when a user
has clicked on an Element so that the system may respond, we must know which one
is being hovered over.

@@

Comparing its type to the mentally convenient type of \texttt{lift hoverable},
we see it is identical except that the Boolean value is a signal, not a signal
of signals. That is, it has been joined.\footnote{While the name
    \texttt{hoverableJoin} reinforces this fact for our paper, we do not
recommend it for a user-facing API.} It is implemented using the more general
primitive \texttt{hoverables} (note the plural), of the following type:

@@

The polymorphic \texttt{a} type can serve as an identifier. The first
value supplies the default value of \texttt{events} (signals must always
be defined and so a default value is required). The returned record
includes the \texttt{events} signal and the \texttt{hoverable}
function, which in general may be applied multiple times so that
multiple elements report their hover information on \texttt{events}. The
\texttt{(Bool -> a)} is used to identify which Element
experienced the event; a common use is \texttt{a = (Int, Bool)} where
the integer identifies the Element and the Bool is the event.
Additionally, \texttt{hoverables} is used to implement
\texttt{hoverable}:

@@

It ignores the polymorphism (\texttt{a = Bool}) and instead create a
Boolean signal that is originally false and use the identity function to
not alter the hoverable information. With a simple change, we can create
a function that acts on \texttt{Signal Element} instead of
\texttt{Element}:

@@

More dangerously, it becomes easy to create an infinite loop. Suppose an
Element shrinks on hover. Suppose the cursor hovers on the Element,
which is then replaced by a smaller Element, so the cursor is no longer
hovering on the Element. Then the original Element is put back, but is
now being hovered on! This condition manifests itself as flickering
between the two Elements. It is unavoidable in any language or system
that allows hover targets to change size in response to hover events.

@@

If we wish to create a button whose string label varies, the types allow
the same exact same technique to be applied. (We have not tested whether
this approach works, but the types allow it.) The same goes for text
fields, which can be given text based on what else is known in the
program, and checkboxes. As currently implemented, checkboxes have
Boolean state so dynamic behavior has limited use. However, we can
imagine the ability to dynamically disable (``gray-out'') a checkbox in response
to other choices in a form.

@@

Therefore, a menu is specified by a value of type \texttt{Tree (Signal String)}.
We can lift and map the combinator \texttt{plainText : String -> Element} to
obtain from a menu specification a value of type \texttt{Tree (Signal Element)}.
From there, we can use \texttt{hoverableJoin} to get the hover information of
each Element and create a value of type \texttt{Tree (Signal Element, Signal
Bool)} to represent our menus. \emph{Now} we convert (by structural recursion)
our tree of signals to a signal of trees, a value of type \texttt{Signal (Tree
(Element, Bool))}. We do the rest of our work with pure functions lifted to
operate on the tree, for example, our rendering function \texttt{renderMenu :
    Tree (Element, Bool) -> Element}. Similarly to limiting the use
    of the IO monad in Haskell code, it is advantageous to limit the use of
    signals in Elm code.

Because \texttt{hoverableJoin} was applied throughout the entire tree, it
contains every Element in every menu, and Booleans that indicate which ones to
display. Rather than omitting Elements associated with false, we replace these
hidden submenus with a spacer. A spacer is simply a blank rectangle with a
specified width and height (\texttt{spacer : Int -> Int
-> Element}). For our purposes, the spacer has the same width as
the Element, but a height of 1.\footnote{We were concerned a height of zero
would cause browser rendering issues.} We then switch between menu and the
spacer based on hover information. The spacer helps to align the submenu with
its parent element in a way a \texttt{Maybe Element} would not have done. By
replacing hidden elements with spacers, we are able to use the \texttt{flow}
primitive to position our elements instead of manually doing ``absolute''
positioning.

@@

Applying \texttt{delayFalse} makes a \texttt{Signal Bool} wait a millisecond to
transition from true to false. Unlike a typical delay, false-to-true transitions
still occur immediately. By applying this function to every hovering Boolean in
the menu structure, we give the mouse time to move from a menu Element to its
submenu before that submenu disappears.

@@

Interestingly, we were able to implement menus without explicitly storing the
state of the menu, i.e. without using \texttt{foldp} (or Elm's implementation of
Arrowized FRP known as \texttt{Automaton}). Rather we use the current browser
DOM state to remember state. Instead of storing the menu currently being
displayed on the screen, we can derive it based on which Element (if any) the
mouse is currently hovering over. The elimination of state from our code allows
it to be much cleaner.

@@

Notice the similarity with \texttt{hoverables}. Each call of
\texttt{customButton} provides the identifier event when the button is
clicked, and three (pure) Elements to display: one normally, one on
hover, and one on click. The result is an \texttt{Element}, not a
\texttt{Signal Element}, that nevertheless changes among those three in
response to the mouse. This is possible because the result Element's
dimensions are taken to be the maximum of the three inputs' dimensions.
Even if the Elements have different sizes, the resulting Element and
therefore the hover surface remains fixed in size. Although the same
join technique can be applied, we find it likely that it would not work
as intended.

In the case of TodoFRP, these three Elements are different colors of the ``x''
and the same for each TODO entry. The polymorphic \texttt{a}s are unique
identifiers (ascending integers) for each entry.

The TODO label Elements are dynamic and do not detect hover information.
The button Elements that do detect hover information are known
statically. With \texttt{hoverableJoin}, it becomes possible to
dismiss a TODO by clicking on its text label.

>>>>

We can print the current area to the screen with \texttt{main = lift asText
area}. The primitive \texttt{asText : a -> Element} renders almost anything into
an \texttt{Element}, which represents a DOM element.\footnote{Those following
    along in an Elm compiler, such as the one available at
    \texttt{elm-lang.org/try}, should add \texttt{import Window} to the top of
the file.}

@@

Elm's \texttt{Graphics.Input} library provides a function to obtain the hover
information from an element.

@@

The returned element is visually identical to the argument, but now detects
hover information. The signal of booleans reflects the hover status of
the this element, not the original. This function works well when the
element is pure (not a signal). For example, if the top-level menu items are
known statically and never change, they can be rendered as pure elements with
this function. We can also create a function, perhaps of type \texttt{Bool ->
Element}, which we can lift onto the hover data to display the submenu only on
mouseover. The result is a value of type \texttt{Signal Element}. If we try to
lift \texttt{hoverable} to accept this value, we get

@@

This is rather unfortunate, as we need to detect hover information on a dynamic
element for a number of reasons. First, after a menu is extended by hovering
over its parent, it must remain so as long as the mouse hovers over the menu
itself. Without this information, the menu will disappear as soon as we try to
mouse over it. Secondly, if we wish the indicate the currently hovered-over item
to the user by visually highlighting it, the element must be dynamic. Thirdly,
we do not wish to require that menu labels be string literals in the source
code, which would be necessary if we could not use signals of elements. Rather,
they may be retrieved or generated in some other manner, or they may change as
the user switches among applications. Finally, in order to determine when a user
has clicked on an element so that the system may respond, we must know which one
is being hovered over.

@@

Comparing its type to the mentally convenient type of \texttt{lift hoverable},
we see it is identical except that the boolean value is a signal, not a signal
of signals. That is, it has been joined.\footnote{While the name
    \texttt{hoverableJoin} reinforces this fact for our paper, we do not
recommend it for a user-facing API.} It is implemented using the more general
primitive \texttt{hoverables} (note the plural), of the following type:

@@

The polymorphic \texttt{a} type can serve as an identifier. The first
value supplies the default value of \texttt{events} (signals must always
be defined and so a default value is required). The returned record
includes the \texttt{events} signal and the \texttt{hoverable}
function, which in general may be applied multiple times so that
multiple elements report their hover information on \texttt{events}. The
\texttt{(Bool -> a)} is used to identify which element
experienced the event; a common use is \texttt{a = (Int, Bool)} where
the integer identifies the element and the boolean is the event.
Additionally, \texttt{hoverables} is used to implement
\texttt{hoverable}:

@@

It ignores the polymorphism (\texttt{a = Bool}) and instead create a
boolean signal that is originally false and use the identity function to
not alter the hoverable information. With a simple change, we can create
a function that acts on \texttt{Signal Element} instead of
\texttt{Element}:

@@

More dangerously, it becomes easy to create an infinite loop. Suppose an
element shrinks on hover. Suppose the cursor hovers on the element,
which is then replaced by a smaller element, so the cursor is no longer
hovering on the element. Then the original element is put back, but is
now being hovered on! This condition manifests itself as flickering
between the two elements. It is unavoidable in any language or system
that allows hover targets to change size in response to hover events.

@@

If we wish to create a button whose string label varies, the types allow
the same exact same technique to be applied. (We have not tested whether
this approach works, but the types allow it.) The same goes for text
fields, which can be given text based on what else is known in the
program, and checkboxes. As currently implemented, checkboxes have
boolean state so dynamic behavior has limited use. However, we can
imagine the ability to dynamically disable (``gray-out'') a checkbox in response
to other choices in a form.

@@

Therefore, a menu is specified by a value of type \texttt{Tree (Signal String)}.
We can lift and map the combinator \texttt{plainText : String -> Element} to
obtain from a menu specification a value of type \texttt{Tree (Signal Element)}.
From there, we can use \texttt{hoverableJoin} to get the hover information of
each element and create a value of type \texttt{Tree (Signal Element, Signal
Bool)} to represent our menus. \emph{Now} we convert (by structural recursion)
our tree of signals to a signal of trees, a value of type \texttt{Signal (Tree
(Element, Bool))}. We do the rest of our work with pure functions lifted to
operate on the tree, for example, our rendering function \texttt{renderMenu :
    Tree (Element, Bool) -> Element}. Similarly to limiting the use
    of the IO monad in Haskell code, it is advantageous to limit the use of
    signals in Elm code.

Because \texttt{hoverableJoin} was applied throughout the entire tree, it
contains every element in every menu, and booleans that indicate which ones to
display. Rather than omitting elements associated with false, we replace these
hidden submenus with a spacer. A spacer is simply a blank rectangle with a
specified width and height (\texttt{spacer : Int -> Int
-> Element}). For our purposes, the spacer has the same width as
the element, but a height of 1.\footnote{We were concerned a height of zero
would cause browser rendering issues.} We then switch between menu and the
spacer based on hover information. The spacer helps to align the submenu with
its parent element in a way a \texttt{Maybe Element} would not have done. By
replacing hidden elements with spacers, we are able to use the \texttt{flow}
primitive to position our elements instead of manually doing ``absolute''
positioning.

@@

Applying \texttt{delayFalse} makes a \texttt{Signal Bool} wait a millisecond to
transition from true to false. Unlike a typical delay, false-to-true transitions
still occur immediately. By applying this function to every hovering boolean in
the menu structure, we give the mouse time to move from a menu element to its
submenu before that submenu disappears.

@@

Interestingly, we were able to implement menus without explicitly storing the
state of the menu, i.e. without using \texttt{foldp} (or Elm's implementation of
Arrowized FRP known as \texttt{Automaton}). Rather we use the current browser
DOM state to remember state. Instead of storing the menu currently being
displayed on the screen, we can derive it based on which element (if any) the
mouse is currently hovering over. The elimination of state from our code allows
it to be much cleaner.

@@

Notice the similarity with \texttt{hoverables}. Each call of
\texttt{customButton} provides the identifier event when the button is
clicked, and three (pure) elements to display: one normally, one on
hover, and one on click. The result is an \texttt{Element}, not a
\texttt{Signal Element}, that nevertheless changes among those three in
response to the mouse. This is possible because the result element's
dimensions are taken to be the maximum of the three inputs' dimensions.
Even if the elements have different sizes, the resulting element and
therefore the hover surface remains fixed in size. Although the same
join technique can be applied, we find it likely that it would not work
as intended.

In the case of TodoFRP, these three elements are different colors of the ``x''
and the same for each TODO entry. The polymorphic \texttt{a}s are unique
identifiers (ascending integers) for each entry.

The TODO label elements are dynamic and do not detect hover information.
The button elements that do detect hover information are known
statically. With \texttt{hoverableJoin}, it becomes possible to
dismiss a TODO by clicking on its text label.
